Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-09-21 analysis/paper/paper.qmd
editing up to discussion section

title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Spheroids are widely distributed lithic artifacts, yet their production strategies and functions remain debated. Traditional morphology-based typologies obscure their technological features and variability. Here we introduce a novel, open-source and reproducible framework that uses spherical harmonics as a representative method of quantitative morphological analysis, integrated with technological measures, to assess production strategies. Applied to Middle Pleistocene faceted spheroids from Qianshangying (~429 ka, northern China), the approach demonstrates the spheroid production followed a conceptual template, involving a patterned reduction process aimed at achieving standardized spherical forms, and represents a deliberate shaping strategy. Our results further demonstrate that spheroids can be clearly distinguished from polyhedrons and multifacial cores exhausted through reduction. The methods for investigating spheroid production strategies reveal a higher level of technological complexity than traditionally assumed for early lithic industries in northern China, and, more importantly, provide a transparent and replicable framework with broad potential for advancing cross-regional and comparative studies of spheroid production and broader lithic technological research.
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: console
---

@@

Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range, from the Oldowan and Acheulean to the Middle Paleolithic, and in diverse archaeological assemblages[@willoughby1985; @Cabanes2024]. Initially, @clark1955 was the pioneer to conduct the study of archaeological stone balls. Then @kleindienst1962 proposed the description of missile, polyhedron and bola. After that, the terminology of spheroid and subspheroid was formally introduced by @leakey1971 in his classification of the Developed Oldowan. At present, commonly used definitions of ball-like stone artifacts are as follows: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex. Compared to spheroids, polyhedrons are characterized by multiple intersecting flake removals, producing a distinctly angular morphology; bolas exhibit smoother and more rounded surfaces with no visible crests [@leakey1971; @tittonSubspheroidsLithicAssemblage2020; @deweyer2017].

@@

This debate remains stalled for several reasons. Typological studies relied heavily on researchers' observations and experiential judgments, which lack objectivity. Some experiments attempted to demonstrate whether spheroids were intentionally or unintentionally produced, yet the researchers involved tend to introduce bias, making it impossible to achieve truly “unconscious” production. In addition, from an objective perspective, most studies can only reflect the characteristics of spheroids from a specific region or site, whereas spheroids themselves display considerable diversity in aspects such as raw materials and surface features. Also, conclusions drawn from their own analyses are rarely subject to external validation or replication. As a result, both the research methods and the inherent diversity of spheroids have hindered the advancement of in-depth studies on spheroids.

To address these challenges, we developed a set of quantitative methods that integrate morphological and technological analyses to investigate spheroid production strategies. While previous morphological measures have focused on the shape of spheroids close to perfect sphere, contributing to discussions of spheroid standardization, they provided limited assistance in reflecting on production strategies. We therefore adopted spherical harmonics, a landmark-free method recently introduced to archaeology. This offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks. We introduce a novel modification of the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons. To our knowledge, this is the first attempt of its kind in the field of archaeology. Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production process. All of the methods we present here are open-source to enable replication of our results and support unified, cross-regional quantitative studies in the future <!-- maybe scar orientation analysis is not open-source?-->.

We demonstrate these methods in a case study of spheroids recovered from the recently excavated site of Qianshangying in the Nihewan Basin of Northern China. In the following sections, we assess whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology, if so, the conceptual template and specialized production intention can be proven. We adopt a comparative approach, comparing spheroids with polyhedrons and multifacial cores. Our comparative study serves two purposes: first, to test whether spheroids follow a distinct production sequence, and second, to explore the potential, beyond typological approaches, of using quantitative methods to distinguish spheroids from cores subjected to intensive reduction. Further, we situated the Middle Pleistocene Qianshangying site within the broader context of East Asia to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the first contributions to understanding the spatial and temporal distribution of spheroids in this region. Most importantly, we offer an efficient and transparent framework combining morphological and technological analyses, enabling comparative studies of spheroid production strategies across regions and archaeological sites.

@@

```{r, echo=FALSE}
#| label: fig-QSY-geography-chronology
#| fig-cap: "Geological Context and Chrono-Stratigraphy of the QSY Site)"

@@

```{r}
#| label: fig-basic-characteristics
#| fig-cap: "Basic characteristics of the Qianshangying assemblage. A: Raw material proportions. B: Box plot comparing of core, spheroid, and multifacial core dimensions"
#| echo: false

@@

# Method

We generated 3D models of the cores for morphometric and technological analysis. Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio 18. The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/xxxx/.

@@

Our work combines the merits of these previous studies to demonstrate an open-source and reproducible analysis of stone artefacts using SPHARM. We further introduce the use of rotation-invariant power spectrum and UMAP to extract shape features, explore hidden structures in data, and then achieve grouping among spheroids and cores, marking a first in archaeological research while significantly extending the depth of analysis provided by SPHARM. All these steps were completed in Python. We hope to demonstrate the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.

```{r, echo=FALSE}
#| label: fig-SPHARM-steps
#| fig-cap: "Schematic summary of the key steps of 3D spherical harmonics analysis. a. UMAP on SPHARM power spectrum for the dimensionality reduction and data structure analysis; b. Spherical harmonics energy (SHE) derived from the SPHARM power spectrum to evaluate the similarity between shapes and the perfect sphere"

@@

Additionally, we computed the spherical harmonic energy (SHE), defined as the sum of squared normalized spherical harmonic coefficients for each degree. This provides a rotation-invariant measure of overall shape complexity and surface regularity. For normalized shapes, this total energy approaches a value of one for a perfect sphere, while more irregular shapes yield higher values.

Notably, classic sphericity metrics [@Wadell1935] exhibit critical limitations in differentiating morphotypes with equivalent surface-area-to-volume ratios. SHE overcomes this constraint through spherical harmonic decomposition of surface morphology, and this allows it to distinguish localized irregularities (like core edge or concave) from global shape deviations (e.g., elongation)-differences that classic sphericity metrics often fail to capture.

@@

We measured spheroid shape with four metrics: sphericity, and spherical harmonic energy (SHE), along with mean edge angle, mean surface curvature. To assess mean edge angle, ten edge intersections were randomly sampled on each spheroid, and edge angles were calculated following @tittonSubspheroidsLithicAssemblage2020by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computing the angles from these coordinates using trigonometric functions in R. Higher mean edge angles indicate a more spherical geometry. We estimated local curvature for each triangular face of the 3D mesh model using a k-nearest neighbors approach in Python. For each triangle, a best-fit plane was fitted to its neighboring triangle centers, and the angle between the triangle’s normal and the best-fit plane’s normal was computed. The average of these angles across all triangles was served as mean surface curvature and quantified overall surface roughness, with larger values indicating greater roughness. Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies-characteristics expected in deliberate shaping strategies.

@@

@@

```{r, echo=FALSE}
#| label: fig-SPHARM-Reconstruction-by-Degree
#| fig-cap: "Shape reconstruction of archaeological spheroids using SPHARM (various lmax levels)"

@tbl-1 summarizes the basic information (rock type, blank type, final morphology, and cortex coverage) and morphological analysis results (classifications derived from SPHARM, values for sphericity and spherical harmonic energy) of the spheroids, subspheroids, polyhedrons and multifical cores involved in the research. From this table, we can observe distinctions between spheroids and subspheroids (n = `r`n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n =`r n_poly_multi`) on the other. In the subsequent analysis, these categories are treated as two separate groups for comparison.

Subsequently, we computed the rotation-invariant power spectrum. Here we present the power spectrum of different groups, including the group of spheroids, group of polyhedrons and multifacial cores, and the total of all these categories combined (@fig-sp-summary). The results show that the spheroid group has significantly lower power than polyhedrons and multifacial cores at both low and high degrees, indicating that spheroids and subspheroids possess a more symmetrical overall shape and more regular local features. In addition, the spheroid group exhibits smaller inter-group morphological variation than polyhedrons and multifacial cores, particularly at low degrees. Then we performed a Variance Analysis on the power of each harmonic degree. The results indicate that l=2 and l=3 exhibit the highest variance across the sample, suggesting that these low-degrees capture the most significant shape variation. Specifically, l=2 reflects elongation or flattening, while l=3 reflects a sphere with octupole (pear-shaped) deformation exhibits a triaxial asymmetry that departs from both spherical and ellipsoidal symmetry.

@@

The spherical harmonic energy (SHE), a novel metric introduced here, quantifies the geometric approximation of a shape to an ideal sphere. Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity. Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy for how closely an object approximates a spherical form (@fig-sp-summary).

Statistical comparisons reveal distinct SHE distributions between groups: the group of spheroids exhibit a mean SHE of `r mean_SHE_sub_spheroid`, whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_SHE_poly_multi`). The Wilcoxon rank-sum test confirms robust inter-group divergence (W = `r U_value2`, p = $`r p_value_Mann_sci2`$), confirming the distinguishability of spheroid group from Qianshangying.

The pursuit of a standardized and distinct spherical form among spheroids is further supported by our findings. Using the Coefficient of Variation (CV), defined as the sample standard deviation divided by the sample mean, we assessed the consistency of sphericity within the spheroid and subspheroids. @Eerkens2001 suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts. Both sphericity and SHE values yielded low CVs (`r round(cv_sph_sub_spheroids, 3)` and `r round(cv_she_sub_spheroids, 3)`, respectively), indicating that group of spheroids maintain a highly standardized rounded shape. In contrast, the group of multifacial cores and polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.

@@

@@

Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 4 pieces in phase one (see supplymentary material-QSY_A.xlsx and QSY_B.xlsx). However, one subspheroid bypassed Phase 1 and instead employed a strategy of bifacial centripetal exploitation combined with edge refinement. Notably, in this case the angle between the two flaking surfaces approaches 90°, distinguishing this subspheroid from typical core of discoid. This indicates that the purpose of reduction was not the production of flakes, but rather the direct shaping of the artifact into its final spherical form. Also, it is important to note that due to the high degree of reduction intensity observed in some spheroids, phases prior to the visible sequence were often obliterated. We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three [@fig-shaping-process].

@@

# scar number ~ SDI
val2 <- labels_long[labels_long$Var1 == "Scar.number" & labels_long$Var2 == "SDI", ]
cor_scar_sdi <- sprintf("%.2f", val2$cor)
p_scar_sdi <- pretty_print_sci(val2$p)

# removal ratio ~ SDI
val3 <- labels_long[labels_long$Var1 == "Removal.ratio" & labels_long$Var2 == "SDI", ]
cor_removal_sdi <- sprintf("%.2f", val3$cor)
p_removal_sdi <- pretty_print_sci(val3$p)

# edge angle ~ sphericity
val4 <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "Sphericity", ]
cor_angle_sph <- sprintf("%.2f", val4$cor)
p_angle_sph <- pretty_print_sci(val4$p)

# edge angle ~ SHE
val5 <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "SHE", ]
cor_angle_she <- sprintf("%.2f", val5$cor)
p_angle_she <- pretty_print_sci(val5$p)

# curvature ~ Sphericity
val6 <- labels_long[labels_long$Var1 == "Curvature" & labels_long$Var2 == "Sphericity", ]
cor_cur_sph <- sprintf("%.2f", val6$cor)
p_cur_sph <- pretty_print_sci(val6$p)

# Edge angle ~ Scar number
val_edge_scar <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "Scar.number", ]
cor_edge_scar <- sprintf("%.2f", val_edge_scar$cor)
p_edge_scar <- pretty_print_sci(val_edge_scar$p)

# Edge angle ~ Removal ratio
val_edge_removal <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "Removal.ratio", ]
cor_edge_removal <- sprintf("%.2f", val_edge_removal$cor)
p_edge_removal <- pretty_print_sci(val_edge_removal$p)

# SHE ~ Removal ratio
val_she_removal <- labels_long[labels_long$Var1 == "SHE" & labels_long$Var2 == "Removal.ratio", ]
cor_she_removal <- sprintf("%.2f", val_she_removal$cor)
p_she_removal <- pretty_print_sci(val_she_removal$p)

# SHE ~ SDI
val_she_sdi <- labels_long[labels_long$Var1 == "SHE" & labels_long$Var2 == "SDI", ]
cor_she_sdi <- sprintf("%.2f", val_she_sdi$cor)
p_she_sdi <- pretty_print_sci(val_she_sdi$p)

>>>>

title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Spheroids are widely distributed lithic artifacts, yet their production strategies and functions remain debated. Traditional morphology-based typologies obscure their technological features and variability. Here we introduce a novel, open-source and reproducible framework that uses spherical harmonics as a method of quantitative morphological analysis, integrated with technological measures, to assess production strategies. Applied to Middle Pleistocene faceted spheroids from Qianshangying (~429 ka, northern China), the approach demonstrates that spheroid production followed a conceptual template, involving a patterned reduction process aimed at achieving standardized spherical forms, and represents a deliberate shaping strategy. Our results further demonstrate that spheroids can be clearly distinguished from polyhedrons and multifacial cores exhausted through reduction. Our methods for investigating spheroid production strategies reveal a higher level of technological complexity than previously assumed for early lithic industries in northern China. Our study also provides a transparent and replicable framework with broad potential for advancing cross-regional and comparative studies of spheroid production but also more broadly lithic technological research for any period and region.
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: console
---

@@

Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range, from the Oldowan and Acheulean to the Middle Paleolithic, and in diverse archaeological assemblages [@willoughby1985; @Cabanes2024]. @clark1955 pioneered the study of archaeological stone balls, then @kleindienst1962 proposed the sub-divisions of missile, polyhedron and bola. After that, the terminology of spheroid and subspheroid was formally introduced by @leakey1971 in his classification of the Developed Oldowan. At present, commonly used definitions of ball-like stone artifacts are as follows: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex. Compared to spheroids, polyhedrons are characterized by multiple intersecting flake removals, producing a distinctly angular morphology; bolas exhibit smoother and more rounded surfaces with no visible crests [@leakey1971; @tittonSubspheroidsLithicAssemblage2020; @deweyer2017].

@@

This debate remains stalled for several reasons. Typological studies relied heavily on researchers' observations and experiential judgments, which lack objectivity. Some experiments attempted to demonstrate whether spheroids were intentionally or unintentionally produced, yet the researchers involved tend to introduce bias, making it impossible to achieve truly “unconscious” production. In addition, from an objective perspective, most studies can only reflect the characteristics of spheroids from a specific region or site, whereas spheroids themselves display considerable diversity in aspects such as raw materials and surface features. Also, conclusions drawn from their own analyses are rarely subjected to external validation or replication. As a result, both the research methods and the inherent diversity of spheroids have hindered the advancement of in-depth studies on spheroids.

To address these challenges, we developed a set of quantitative methods that integrate morphological and technological analyses to investigate spheroid production strategies. While previous morphological measures have focused on the shape of spheroids close to perfect sphere, contributing to discussions of spheroid standardization, they provided limited assistance in reflecting on production strategies. We therefore adopted spherical harmonics, a landmark-free method recently introduced to archaeology. This offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks. We introduce a novel modification of the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons. To our knowledge, this is the first attempt of its kind in the field of archaeology. Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production process. All of the methods we present here are open-source to enable replication of our results and support unified, cross-regional quantitative studies in the future.

We demonstrate these methods in a case study of spheroids recovered from the recently excavated site of Qianshangying in the Nihewan Basin of Northern China. In the following sections, we assess whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology. If these can be demonstrated, the conceptual template and specialized production intention can be supported. We adopt a comparative approach, comparing spheroids with polyhedrons and multifacial cores. Our comparative study serves two purposes: first, to test whether spheroids follow a distinct production sequence, and second, to explore the potential, beyond typological approaches, of using quantitative methods to distinguish spheroids from cores subjected to intensive reduction. Further, we situated the Middle Pleistocene Qianshangying site within the broader context of East Asia to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the first contributions to understanding the spatial and temporal distribution of spheroids in this region. Finally, and most generally, we offer an efficient and transparent framework combining morphological and technological analyses, enabling comparative studies of spheroid production strategies across regions and archaeological sites that can also be extended to other types of artefacts.

@@

```{r, echo=FALSE}
#| label: fig-QSY-geography-chronology
#| fig-cap: "Geological Context and Chrono-Stratigraphy of the Qianshangying site complex)"

@@

```{r}
#| label: fig-basic-characteristics
#| fig-cap: "Basic characteristics of the Qianshangying assemblage. A: Raw material proportions. B: Box plot comparing basic metric dimensions of cores, spheroids, and multifacial cores"
#| echo: false

@@

# Methods

We generated 3D models of the cores for morphometric and technological analysis. Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio 18. The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://osf.io/ctne9/.

@@

Our work combines the merits of these previous studies to demonstrate an open-source and reproducible analysis of stone artefacts using SPHARM. We further introduce the use of rotation-invariant power spectrum and UMAP to extract shape features, explore hidden structures in data, and then achieve grouping among spheroids and cores, marking a first in archaeological research while substantially extending the depth of analysis provided by SPHARM. All these steps were completed in Python with the code openly available online at https://osf.io/ctne9/. Our work demonstrates the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.

```{r, echo=FALSE}
#| label: fig-SPHARM-steps
#| fig-cap: "Schematic summary of the key steps of 3D spherical harmonics analysis. See text for more details. a. UMAP on SPHARM power spectrum for the dimensionality reduction and data structure analysis; b. Spherical harmonics energy (SHE) derived from the SPHARM power spectrum to evaluate the similarity between shapes and the perfect sphere"

@@

We computed the spherical harmonic energy (SHE), defined as the sum of squared normalized spherical harmonic coefficients for each degree. This provides a rotation-invariant measure of overall shape complexity and surface regularity. For normalized shapes, this total energy approaches a value of one for a perfect sphere, while more irregular shapes yield higher values.

At noted above, classic sphericity metrics [@Wadell1935] exhibit critical limitations in differentiating morphotypes with equivalent surface-area-to-volume ratios. SHE overcomes this constraint through spherical harmonic decomposition of surface morphology, and this allows it to distinguish localized irregularities (like core edge or concave) from global shape deviations (e.g., elongation)-differences that classic sphericity metrics often fail to capture.

@@

We measured spheroid shape with four metrics: sphericity, and spherical harmonic energy (SHE), along with mean edge angle, mean surface curvature. To assess mean edge angle, ten edge intersections were randomly sampled on each spheroid, and edge angles were calculated following @tittonSubspheroidsLithicAssemblage2020 by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computing the angles from these coordinates using trigonometric functions in R. Higher mean edge angles indicate a more spherical geometry. We estimated local curvature for each triangular face of the 3D mesh model using a k-nearest neighbors approach in Python. For each triangle, a best-fit plane was fitted to its neighboring triangle centers, and the angle between the triangle’s normal and the best-fit plane’s normal was computed. The average of these angles across all triangles was served as mean surface curvature and quantified overall surface roughness, with larger values indicating greater roughness. Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies-characteristics expected in deliberate shaping strategies.

@@

@tbl-1 summarizes the basic information (rock type, blank type, final morphology, and cortex coverage) and morphological analysis results (classifications derived from SPHARM, values for sphericity and spherical harmonic energy) of the spheroids, subspheroids, polyhedrons and multifical cores involved in the research. From this table, we can observe distinctions between spheroids and subspheroids (n = `r n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n =`r n_poly_multi`) on the other. In the subsequent analysis, these categories are treated as two separate groups for comparison.

```{r, echo=FALSE}
#| label: fig-SPHARM-Reconstruction-by-Degree
#| fig-cap: "Shape reconstruction of selected archaeological spheroids using SPHARM (various lmax levels)"

Subsequently, we computed the rotation-invariant power spectrum. Here we present the power spectrum of different groups, including the group of spheroids, group of polyhedrons and multifacial cores, and the total of all these categories combined (@fig-sp-summary). The results show that the spheroid group has significantly lower power than polyhedrons and multifacial cores at both low and high degrees (panel A @fig-sp-summary), indicating that spheroids and subspheroids possess a more symmetrical overall shape and more regular local features. In addition, the spheroid group exhibits smaller inter-group morphological variation than polyhedrons and multifacial cores, particularly at low degrees. Then we performed a Variance Analysis on the power of each harmonic degree. The results indicate that l=2 and l=3 exhibit the highest variance across the sample (panel B @fig-sp-summary), suggesting that these low-degrees capture the most significant shape variation. Specifically, l=2 reflects elongation or flattening, while l=3 reflects a sphere with octupole (pear-shaped) deformation exhibits a triaxial asymmetry that departs from both spherical and ellipsoidal symmetry.

Statistical comparisons reveal distinct SHE distributions between groups: the group of spheroids exhibit a mean SHE of `r mean_SHE_sub_spheroid`, whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_SHE_poly_multi`) (panel C @fig-sp-summary). The Wilcoxon rank-sum test confirms robust inter-group divergence (W = `r U_value2`, p = $`r p_value_Mann_sci2`$), confirming the distinguishability of spheroid group from Qianshangying.

The pursuit of a standardized and distinct spherical form among spheroids is supported by our findings. Using the Coefficient of Variation (CV), defined as the sample standard deviation divided by the sample mean, we assessed the consistency of sphericity within the spheroid and subspheroids. @Eerkens2001 suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts. Both sphericity and SHE values yielded low CVs (`r round(cv_sph_sub_spheroids, 3)` and `r round(cv_she_sub_spheroids, 3)`, respectively), indicating that group of spheroids maintain a highly standardized rounded shape (panel C @fig-sp-summary). In contrast, the group of multifacial cores and polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.

The spherical harmonic energy (SHE), a novel metric introduced here, quantifies the geometric approximation of a shape to an ideal sphere. Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity. Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy for how closely an object approximates a spherical form (panel D @fig-sp-summary).

@@

@@

@@

Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 4 pieces in phase one (see supplementary material files QSY_A.xlsx and QSY_B.xlsx). However, one subspheroid bypassed Phase 1 and instead employed a strategy of bifacial centripetal exploitation combined with edge refinement. Notably, in this case the angle between the two flaking surfaces approaches 90°, distinguishing this subspheroid from typical core of discoid. This indicates that the purpose of reduction was not the production of flakes, but rather the direct shaping of the artifact into its final spherical form. Also, it is important to note that due to the high degree of reduction intensity observed in some spheroids, phases prior to the visible sequence were often obliterated. We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three [@fig-shaping-process].

@@

@@

# scar number ~ SDI
val2 <- labels_long[labels_long$Var1 == "Scar.number" & labels_long$Var2 == "SDI", ]
cor_scar_sdi <- sprintf("%.2f", val2$cor)
p_scar_sdi <- sprintf("%.3f", val2$p)

# removal ratio ~ SDI
val3 <- labels_long[labels_long$Var1 == "Removal.ratio" & labels_long$Var2 == "SDI", ]
cor_removal_sdi <- sprintf("%.2f", val3$cor)
p_removal_sdi <- sprintf("%.3f", val3$p)

# edge angle ~ sphericity
val4 <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "Sphericity", ]
cor_angle_sph <- sprintf("%.2f", val4$cor)
p_angle_sph <- sprintf("%.3f", val4$p)

# edge angle ~ SHE
val5 <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "SHE", ]
cor_angle_she <- sprintf("%.2f", val5$cor)
p_angle_she <- sprintf("%.3f", val5$p)

# curvature ~ Sphericity
val6 <- labels_long[labels_long$Var1 == "Curvature" & labels_long$Var2 == "Sphericity", ]
cor_cur_sph <- sprintf("%.2f", val6$cor)
p_cur_sph <- sprintf("%.3f", val6$p)

# Edge angle ~ Scar number
val_edge_scar <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "Scar.number", ]
cor_edge_scar <- sprintf("%.2f", val_edge_scar$cor)
p_edge_scar <- sprintf("%.3f", val_edge_scar$p)

# Edge angle ~ Removal ratio
val_edge_removal <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "Removal.ratio", ]
cor_edge_removal <- sprintf("%.2f", val_edge_removal$cor)
p_edge_removal <- sprintf("%.3f", val_edge_removal$p)

# SHE ~ Removal ratio
val_she_removal <- labels_long[labels_long$Var1 == "SHE" & labels_long$Var2 == "Removal.ratio", ]
cor_she_removal <- sprintf("%.2f", val_she_removal$cor)
p_she_removal <- sprintf("%.3f", val_she_removal$p)

# SHE ~ SDI
val_she_sdi <- labels_long[labels_long$Var1 == "SHE" & labels_long$Var2 == "SDI", ]
cor_she_sdi <- sprintf("%.2f", val_she_sdi$cor)
p_she_sdi <- sprintf("%.3f", val_she_sdi$p)
